---
title: "RNA-seq Analysis of Ism1-/- Mouse Lung Tissue (GSE262122)"
author: "Mario Antonio Rodriguez Diaz"
date: "2026-03-21"
output:
  pdf_document: default
  html_document: default
---


```{r setup, include=FALSE}
# rnaseq_deseq2_ism1_lung.Rmd
# RNA-seq differential expression analysis: Ism1-/- vs Wild Type (GSE262122)
# Dataset: Mus musculus lung tissue, postnatal day 7, Illumina NovaSeq 6000
#
# Author: Mario Antonio Rodriguez Diaz
# MSc Bioinformatics, University of Edinburgh, 2025-2026
# Functional Genomic Technologies coursework
# Adapted from class template (Tomlinson, University of Edinburgh)

knitr::opts_chunk$set(echo = TRUE)
```


```{r init, echo=FALSE}
# Code adapted from class template: example_analysis_GSE243520.Rmd
# (Tomlinson, University of Edinburgh, 2026)
#Load the required libraries & files
library(Rsubread)
library(edgeR)
library(limma)
library(gplots)
library(DESeq2)
library(affy)
library(QuasR)
library(pheatmap)

load("mytable_feaures")

```

Once we have loaded the data we are ready to run some further analysis.
```{r input_processing, echo=FALSE}
#Some further processing to access count data and column names.
table_rnaseq<-(mytable_feaures)$counts 
colnames_rnaseq <-colnames(table_rnaseq)

#Here I generate an 'adf data frame' for the annotation.  
adf <-data.frame(Colnames =colnames_rnaseq)
library(stringr)
locc <- str_locate_all(pattern="Aligned.sortedByCoord.out.bam",adf$Colnames)
adf$ShortGEOName <-substr(adf$Colnames,1,locc[[1]][,"start"]-1)

#From the GEO page for this study-https://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE262122
#GSM8157908 	WT_1
#GSM8157909 	WT_2
#GSM8157910 	WT_3
#GSM8157911 	Ism1_1
#GSM8157912 	Ism1_2
#GSM8157913 	Ism1_3

#My samples are in the same order as the annotation so I can just create some short names and add to annotation
adf$ShortNames <-c("WT_1", "WT_2", "WT_3", "Ism1_1", "Ism1_2", "Ism1_3")

#while I am doing this I will create a factor for the genotype
split_names <-str_split(adf$ShortNames, pattern="_")
genotype <-c()
for(i in 1:length(split_names)){
  genotype <-c(genotype,split_names[[i]][1])
}
# Creating new Genotype col in adf as a factor, specifying that "WT" is the control
adf$Genotype <- factor(genotype,levels = c("WT","Ism1")) 

#Add some colours as well
#See https://stackoverflow.com/questions/42288075/how-to-autogenerate-color-vector-from-a-numeric-vector
z <- adf$Genotype # no need to specify that Z is a factor because col Genotype was created as a factor
adf$Colours <-rainbow(length(levels(z)))[z]


#make sure the colnames of the study match the rownames of the annotation...
rownames(adf) <-adf$Colnames
if(identical(colnames_rnaseq,rownames(adf))==FALSE){
  print("Names do not match")
}else{
  print("Names match.. OK to proceed")
}

#OK switch to shortnames to help with the figures
#We can only swap this way because column order to consistent with adf
rownames(adf) <-adf$ShortNames
colnames(table_rnaseq) <-adf$ShortNames

colnames_rnaseq <-colnames(table_rnaseq)

```

## Loading the Data into DESeq
DESeq2 will used for the statistical analysis and the normalization of the data. 
```{r loading_deseq, echo=FALSE}
#Create object for DESeq
#NOTE the design for differential analysis is fed into this object 
dds_rnaseq <- DESeqDataSetFromMatrix(countData = table_rnaseq,colData = adf,design = ~ Genotype)
#check the object dimensions
dim(dds_rnaseq)
head(rownames(dds_rnaseq))
colnames(dds_rnaseq)
#check the object type
class(dds_rnaseq)
typeof(dds_rnaseq)

#check what this object contains
slotNames(dds_rnaseq)
colnames(colData(dds_rnaseq))
colnames(assay(dds_rnaseq))

```
Quality Filtering Genes
Apply filters to remove genes that do not pass quality filters.

```{r apply_filters, echo=FALSE}
#first filter based upon counts
keep <- rowSums(counts(dds_rnaseq)) > 1
dds_rnaseq <- dds_rnaseq[keep,]
nrow(dds_rnaseq)
```

## Normalisation of Data

```{r normalisation, echo=TRUE}
#size factor estimation for normalisation- these are used to scale by library size when data is exported.
dds_rnaseq <- estimateSizeFactors(dds_rnaseq)
sizeFactors(dds_rnaseq)

```

We will use rlog normalisation
```{r plot_normalised, echo=FALSE}

rld_rnaseq <- rlog(dds_rnaseq, blind = T)
head(rld_rnaseq, 3)


```

Plotting normalisation
```{r build_plots, echo=FALSE}
#Splitting screen in 1 row, 2 cols
par(mfrow = c(1, 2))

#Plotting rlog normalised data
boxplot(assay(rld_rnaseq),
        col = adf$Colours, 
        main = "Normalised data (rlog)",
        ylab = "rlog Counts",
        las = 2)

#Plotting non-normalised data
counts_un <-counts(dds_rnaseq, normalized=FALSE)+1
boxplot(log2(counts_un),
        col = adf$Colours, 
        main = "Non-normalised data (Log2)",
        ylab = "log2(counts + 1)",
        las = 2)
```

Exploratory Data Plots
```{r more_explore, echo=FALSE}
#we can run a PCA plot here...
library(scatterplot3d)
library(ggplot2)
# Perform PCA
pca <- prcomp(t(na.omit(assay(rld_rnaseq))), scale=T)
# Plot the PCA results
s3d<-scatterplot3d(pca$x[,1:3], pch=19, color=adf$Colours,
                   main="3D Principal Component Analysis")
s3d.coords <- s3d$xyz.convert(pca$x[,1:3])
text(s3d.coords$x, s3d.coords$y, labels = colnames(rld_rnaseq),pos = 3,offset = 0.5,cex=0.5)
#or a quick plot in 2d
qplot(pca$x[,1],pca$x[,2],xlab="PCA1", ylab="PCA2",color=adf$Genotype,
      main="PCA: Ism1 Mutant vs Wild Type")
#We can also output a distance matrix…
sampleDists <- dist(t(assay(rld_rnaseq)))
sampleDists

```
## Generate further heatmaps with the data

```{r more_heatmaps, echo=FALSE}
library("pheatmap")
library("RColorBrewer")
sampleDistMatrix <- as.matrix(sampleDists)
pheatmap(sampleDistMatrix, main = "Sample-to-Sample Distance Heatmap")
#easier PCA plot
ress <-plotPCA(rld_rnaseq, intgroup = c("Genotype"))
ress <- ress + geom_label(aes(label = name))+
  ggtitle("PCA Cluster Map: Genotype Comparison")
plot(ress)
```
## Differential Gene Expression
Simplified stats

```{r differential_expression, echo=FALSE}
dds_rnaseq <- DESeq(dds_rnaseq)

#This is how you specify a contrast…
result_treatment <-results(dds_rnaseq,contrast=c("Genotype","Ism1","WT"))
#Checking range of L2FC to further MA plot
range(result_treatment$log2FoldChange)
#make a nice plot of differential expression
plotMA(result_treatment, main="MA Plot: Ism1-/- vs WT", ylim=c(-6.5,6.5), alpha = 0.05)
#Counting number of results for different thresholds
table(result_treatment$pvalue < 0.05)
table(result_treatment$padj < 0.05)
#Ranking result_treatment by the absolute value of stat (decreasing)
#Genes with the largest change (both up-regulated and down-regulated) are ranked first
result_treatment <-result_treatment[order(abs(result_treatment$stat),decreasing = TRUE), ]
head(result_treatment)

#Saving results as csv
write.csv(as.data.frame(result_treatment),"result_treatment_sorted_not_ann.csv")

```

## Make heatmaps for top70 out of best results

```{r differential_expression_heatmaps, echo=FALSE}

top70 <-rownames(result_treatment)[1:70]
#make a heatmap for the top 70
pheatmap(assay(rld_rnaseq)[top70,],scale="row",show_rownames=F,main="RLD values, Row Scaled, Top 70")

```
## Build Gene Annotation. A serialised is generated in case Ensembl is unavailable

```{r build_gene_annotation, echo=TRUE}
#Download Ensembl annotation using BiomaRt and rename the samples
library(biomaRt)
if(file.exists("resultAnnot.RData")){

  load("resultAnnot.RData")
  print("Loaded existing Ensembl annotation...")
}else{
  
  #UK ensembl is being updated  we use a USA mirror, "useast.ensembl.org"
  ensembl_host <-"https://www.ensembl.org"
  
  # Requesting fewer attributes so that the connection to Ensemble does not get interrupted 
  mart <- biomaRt::useDataset(dataset = "mmusculus_gene_ensembl",mart    = useMart("ENSEMBL_MART_ENSEMBL",host    = ensembl_host))
  resultAnnot <- biomaRt::getBM(values=rownames(dds_rnaseq),attributes =     c("ensembl_gene_id","external_gene_name","chromosome_name","description"),filters="ensembl_gene_id",mart=mart)
  save(resultAnnot,file="resultAnnot.RData")
}

```

Merge Annotation with Input Data
```{r merge_gene_annotation, echo=FALSE}
#merge with input data
names <- resultAnnot[,1]
resultAnnot <- as.data.frame(resultAnnot)
rownames(resultAnnot) = names
idx<-match(rownames(dds_rnaseq),rownames(resultAnnot))
#make sure annotation is in same order
all(rownames(dds_rnaseq) == rownames(resultAnnot))
grr<-resultAnnot[match(rownames(dds_rnaseq), resultAnnot$ensembl_gene_id),]
all(rownames(dds_rnaseq) == rownames(grr))
resultAnnot <-grr
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#make the nice names
nice_names<-paste(resultAnnot$ensembl_gene_id,resultAnnot$external_gene_name, sep = '_')
resultAnnot$nice_names <-nice_names
head(resultAnnot)
all(rownames(dds_rnaseq) == rownames(resultAnnot))
#check names
# Changed to create heatmap for top40
rld_rnaseq <- rlog(dds_rnaseq, blind = TRUE)#
idx2 <-match(rownames(result_treatment)[1:40],rownames(dds_rnaseq))
plotme <-(rld_rnaseq)[rownames(result_treatment)[1:40],]
rownames(plotme)<-resultAnnot$nice_names[idx2]
#make heatmap with candidate genes
pheatmap(assay(plotme),scale="row",fontsize_col = 5, fontsize_row = 5,cellheight =6, 
         cellwidth =6,treeheight_row = 60, treeheight_col = 20)
```
## Final Results Table
Since we have annotation that matches gene names to candidates we can also annotate the results table from differential expression

```{r merge_gene_annotation2, echo=FALSE}

#this has the results ordered

#before 
head(result_treatment,50)

idx2 <-match(rownames(result_treatment),rownames(resultAnnot))

# Create a temporary vector of names
new_names <- resultAnnot$nice_names[idx2]

# If a name is NA, replace it with the original Ensembl ID from the row names
new_names[is.na(new_names)] <- rownames(result_treatment)[is.na(new_names)]
# -------------------------------------

# 2. Assign the cleaned names back to the results
rownames(result_treatment) <- new_names
#after
print(result_treatment[1:50,])

#Saving results as csv
write.csv(as.data.frame(result_treatment),"result_treatment_sorted_ann.csv")


```

